Linearly Ordered Attribute Grammar scheduling using SAT-solving

### 2015-01-15 paper/paper.tex
Kleine typefout
Remove some 'much' and 'very' instances from paper

There are two main advantages to using a SAT-solver for scheduling:
(1) the scheduling algorithm runs much faster than existing scheduling
algorithms on real-world examples, and (2) by adding extra constraints
we obtain fine-grained control over the resulting schedule, thereby
enabling new scheduling optimisations.

@@

To encode this problem in SAT we represent each edge in the dependency graphs as a Boolean variable, with its value indicating the direction of the edge. For the direct dependencies the value is already set, but for the rest of the variables the SAT-solver may choose the direction of the edge. Ensuring cycle-freeness requires us to encode transitivity with SAT-constraints, which in the straight-forward solution leads to a number of extra constraints cubic in the number of variables. To avoid that problem we make our graphs \emph{chordal} \cite{Dirac:1961}. In a chordal graph every cycle of size $> 3$ contains an edge between two non-adjacent nodes in the cycle. In other words, if there exists a cycle in the graph, there must also exist a cycle of at most three nodes. This allows us to encode cycle freeness much more efficiently by only disallowing cycles of length three. Chordality has been used previously to encode equality logic in SAT using undirected graphs \cite{bryant-chordal}; to our knowledge this is the first application of chordality to express cycle-freeness of directed graphs.

@@

We have run this algorithm on ever output attribute of the top-level nonterminal of all our examples. For our largest grammar, the hardest output to compute took 1 second. So, while a harder optimisation than the previous one, it is very doable in practice.

@@

\section{Discussion and Conclusion} \label{sect:conclusion}
We have explained the difficulties in attribute grammar scheduling and shown how to solve this problem using a SAT-solver. The given approach has been implemented and tested on a full-scale compiler built using attribute grammars. Results show that the algorithm works much faster than other approaches.

@@

One problem with the current approach in contrast to most other scheduling algorithms is the unpredictability. SAT-solvers use certain heuristics to quickly find solutions in many applications, but it can theoretically happen that for a certain attribute grammar the SAT problem that is generated is not suitable for these heuristics. A seemingly innocent change in the attribute grammar definition could therefore theoretically lead to a large increase in compile time. However, we have not encountered this problem and we believe that this situation is very unlikely to happen because of the maturity of the SAT-solvers.

>>>>

There are two main advantages to using a SAT-solver for scheduling:
(1) the scheduling algorithm runs faster than existing scheduling
algorithms on real-world examples, and (2) by adding extra constraints
we obtain fine-grained control over the resulting schedule, thereby
enabling new scheduling optimisations.

@@

To encode this problem in SAT we represent each edge in the dependency graphs as a Boolean variable, with its value indicating the direction of the edge. For the direct dependencies the value is already set, but for the rest of the variables the SAT-solver may choose the direction of the edge. Ensuring cycle-freeness requires us to encode transitivity with SAT-constraints, which in the straight-forward solution leads to a number of extra constraints cubic in the number of variables. To avoid that problem we make our graphs \emph{chordal} \cite{Dirac:1961}. In a chordal graph every cycle of size $> 3$ contains an edge between two non-adjacent nodes in the cycle. In other words, if there exists a cycle in the graph, there must also exist a cycle of at most three nodes. This allows us to encode cycle freeness more efficiently by only disallowing cycles of length three. Chordality has been used previously to encode equality logic in SAT using undirected graphs \cite{bryant-chordal}; to our knowledge this is the first application of chordality to express cycle-freeness of directed graphs.

@@

We have run this algorithm on every output attribute of the top-level nonterminal of all our examples. For our largest grammar, the hardest output to compute took 1 second. So, while a harder optimisation than the previous one, it is very doable in practice.

@@

\section{Discussion and Conclusion} \label{sect:conclusion}
We have explained the difficulties in attribute grammar scheduling and shown how to solve this problem using a SAT-solver. The given approach has been implemented and tested on a full-scale compiler built using attribute grammars. Results show that the algorithm works faster than other approaches.

@@

One problem with the current approach in contrast to most other scheduling algorithms is the unpredictability. SAT-solvers use certain heuristics to quickly find solutions in many applications, but it can theoretically happen that for a certain attribute grammar the SAT problem that is generated is not suitable for these heuristics. A seemingly innocent change in the attribute grammar definition could therefore theoretically lead to a large increase in compile time. However, we have not encountered this problem and we believe that this situation is unlikely to happen because of the maturity of the SAT-solvers.
